---
output:
  bookdown::gitbook:
    lib_dir: "book_assets"
    includes:
      in_header: google_analytics.html
  html_document: default
  pdf_document: default
---
<!--# Working with data-->
# Trabajar con Datos

```{r echo=FALSE}
library(tidyverse)
library(cowplot)
```

<!--## What are data? -->
## ¿Qué son los datos?

<!--The first important point about data is that data *are* -- meaning that the word "data" is plural (though some people disagree with me on this).  You might also wonder how to pronounce "data" -- I say "day-tah", but I know many people who say "dah-tah", and I have been able to remain friends with them in spite of this. Now, if I heard them say "the data is" then that would be a bigger issue...-->
Cuando hablamos de los datos lo hacemos en plural. Si te encuentras buscando información de estadística en inglés y aparece como "data", recuerda que se trata de una palabra que siempre permanece en plural. 


<!--### Qualitative data
Data are composed of *variables*, where a variable reflects a unique measurement or quantity.  Some variables are *qualitative*, meaning that they describe a quality rather than a numeric quantity.  For example, in my stats course I generally give an introductory survey, both to obtain data to use in class and to learn more about the students. One of the questions that I ask is "What is your favorite food?", to which some of the answers have been: blueberries, chocolate, tamales, pasta, pizza, and mango.  Those data are not intrinsically numerical; we could assign numbers to each one (1=blueberries, 2=chocolate, etc), but we would just be using the numbers as labels rather than as real numbers.  This also constrains what we should do with those numbers; for example, it wouldn't make sense to compute the average of those numbers.  However, we will often code qualitative data using numbers in order to make them easier to work with, as you will see later.-->
### Datos Cualitativos

Los datos se componen de *variables*, en donde una variable refleja una medida o cantidad única. Algunas variables son *cualitativas*, lo que significa que describen una cualidad en lugar de una cantidad numérica. Por ejemplo, en mi curso de estadística generalmente doy un cuestionario introductorio, con el propósito de obtener datos que pueda usar en clase y para aprender más sobre los estudiantes. Una de las preguntas que hago es "¿Cuál es tu comida favorita?", a lo cual algunas de las respuestas han sido: arándanos, chocolate, tamales, pasta, pizza y mango. Esos datos no son esencialmente numéricos; podríamos asignarles números a cada uno (1=arándanos, 2=chocolate, etc), pero solamente estaríamos utilizando los números como etiquetas en lugar de números reales. Esto también restringe lo que podríamos hacer con esos números; por ejemplo, no tendría sentido el calcular el promedio de esos números. Sin embargo, a menudo codificaremos datos cualitativos utilizando números para poder trabajar más facilmente con ellos, como verán más adelante en este libro. 

<!--### Quantitative data
More commonly in statistics we will work with *quantitative* data, meaning data that are numerical.  For example, here Table \@ref(tab:WhyTakingClass) shows the results from another question that I ask in my introductory class, which is "Why are you taking this class?"-->
### Datos cuantitativos

Más comunmente en estadística trabajaremos con datos *cuantitativos*, lo cual significa que los datos son numéricos. Por ejemplo, aquí en la Tabla \@ref(tab:WhyTakingClass) muestra los resultados de otra de las preguntas que realizo en mi clase introductoria, la cual es "¿Por qué estás tomando esta clase?"


```{r WhyTakingClass,echo=FALSE}
# create table showing frequency of responses to question about why students
# are taking statistics class

classData=data.frame(why=c("It fulfills a degree plan requirement",
                           "It fulfills a General Education Breadth Requirement",
                           "It is not required but I am interested in the topic",
                           "Other"),
                     nResponses=c(105,32,11,4))

knitr::kable(
  classData, booktabs = TRUE,
  caption = 'Counts of the prevalence of different responses to the question "Why are you taking this class?"',
  col.names=c('Why are you taking this class?',"Number of students")
)
```

<!--Note that the students' answers were qualitative, but we generated a quantitative summary of them by counting how many students gave each response.-->
Nota que las respuestas de los estudiantes fueron cualitativas, pero generamos un resumen cuantitativo de ellos contando cuántos estudiantes respondieron a cada opción.

<!--#### Types of numbers
There are several different types of numbers that we work with in statistics.  It's important to understand these differences, in part because statistical analysis languages like R often distinguish between them.-->
### Tipos de números

Existen varios tipos diferentes de números con los que trabajamos en estadística. Es importante entender estas diferencias, en parte porque los lenguajes para análisis estadístico como R a menudo los distinguen. 

<!--**Binary numbers**. The simplest are binary numbers -- that is, zero or one.  We will often use binary numbers to represent whether something is true or false, or present or absent.  For example, I might ask 10 people if they have ever experienced a migraine headache, recording their answers as "Yes" or "No".  It's often useful to instead use *logical* values, which take the value of either ```TRUE``` or ```FALSE```.  This can be especially useful when we start using programming languages like R to analyze our data, since these languages already understand the concepts of TRUE and FALSE.  In fact, most programming languages treat truth values and binary numbers equivalently. The number 1 is equal to the logical value ```TRUE```, and the number zero is equal to the logical value ``FALSE``.-->
**Números binarios**. Los más simples son los números binarios -- cero ó uno. A menudo usaremos números binarios para representar si algo es verdadero o falso, o presente o ausente. Por ejemplo, puede que le pregunte a 10 personas si alguna vez han tenido dolor de cabeza por migraña, registrando sus respuestas como "Sí" ó "No". En ocasiones es útil usar valores *lógicos*, los cuales toman los valores de ```VERDADERO``` o ```FALSO```. Esto puede ser especialmente útil cuando comenzamos a utilizar lenguajes de programación como R para analizar nuestros datos, ya que, estos lenguajes comprenden los conceptos de VERDADERO y FALSO. De hecho, la mayoría de los lenguajes de programación tratan los valores de verdad y los números binarios de manera equivalente. El número 1 es igual al valor lógico ```VERDADERO```, y el número cero es igual al valor lógico ```FALSO```.   

<!--**Integers**.  Integers are whole numbers with no fractional or decimal part. We most commonly encounter integers when we count things, but they also often occur in psychological measurement.  For example, in my introductory survey I administer a set of questions about attitudes towards statistics (such as "Statistics seems very mysterious to me."), on which the students respond with a number between 1 ("Disagree strongly") and 7 ("Agree strongly").-->  
**Enteros**. Los enteros son números enteros sin fracción o punto decimal. Nos encontramos más comunmente números enteros cuando contamos cosas, pero también ocurren en la medición de aspectos psicológicos. Por ejemplo, en mi cuestionario introductorio administro un set de preguntas sobre actitudes hacia la estadística (tal como "La estadística me parece misteriosa."), para lo cual les estudiantes responden con un número entre 1 ("Muy en desacuerdo") y 7 ("Muy de acuerdo"). 

<!--**Real numbers**.  Most commonly in statistics we work with real numbers, which have a fractional/decimal part.  For example, we might measure someone's weight, which can be measured to an arbitrary level of precision, from kilograms down to micrograms.-->
**Números reales**. En estadística trabajamos más comunmente con números reales, los cuales tienen parte fraccionaria o decimal. Por ejemplo, cuando medimos el peso de alguien, éste puede ser medido a un nivel arbitrario de precisión, desde kilogramos hasta microgramos. 

<!--## Discrete versus continuous measurements
A *discrete* measurement is one that takes one of a finite set of particular values.  These could be qualitative values (for example, different breeds of dogs) or numerical values (for example, how many friends one has on Facebook).  Importantly, there is no middle ground between the measurements; it doesn't make sense to say that one has 33.7 friends.-->
## Mediciones Discretas versus Continuas 

Una medición *discreta* es aquella que toma uno de un conjunto finito de valores particulares. Estos pueden ser valores cualitativos (por ejemplo, diferentes tipos de razas de perros) o valores numéricos (por ejemplo, cuántos amigos tiene une en facebook). Es importante recordar que, no hay punto medio en las medidas; no tiene sentido decir que une tiene 33.7 amigues.

<!--A *continuous* measurement is one that is defined in terms of a real number. It could fall anywhere in a particular range of values, though usually our measurement tools will limit the precision with which we can measure it; for example, a floor scale might measure weight to the nearest kg, even though weight could in theory be measured with much more precision.-->
Una medición *continua* es aquella que es definida en términos de un número real. Puede encontrarse en cualquier parte de un rango particular de valores, aunque usualmente nuestras herramientas de medición pueden limitar la precisión con la que podemos medirla; por ejemplo, una báscula de piso puede medir el peso al kilogramo más cercano, aunque en teoría el peso puede ser medido con mucha mayor precisión.

<!--It is common in statistics courses to go into more detail about different "scales" of measurement, which are discussed in more detail in the Appendix to this chapter.  The most important takeaway from this is that some kinds of statistics don't make sense on some kinds of data.  For example, imagine that we were to collect postal Zip Code data from a number of individuals. Those numbers are represented as integers, but they don't actually refer to a numeric scale; each zip code basically serves as a label for a different region. For this reason, it wouldn't make sense to talk about the average zip code, for example.-->  
En los cursos de estadística es común revisar con más detalle las diferentes "escalas" de medición, las cuales son discutidas con más detalle en el Apéndice de este capítulo. El punto más importante a recordar de esto es que algunos tipos de estadística no hacen sentido con algunos tipos de datos. Por ejemplo, imagina que reunieramos el código postal de un grupo de individuos. Esos números son representados como enteros, pero en realidad no se refieren a una escala numérica; cada código postal sirve básicamente como etiqueta para una región diferente. Por esta razón, no tendría sentido hablar del código postal promedio.

<!--## What makes a good measurement?-->
## ¿Qué constituye a una buena medición?

<!--In many fields such as psychology, the thing that we are measuring is not a physical feature, but instead is an unobservable theoretical concept, which we usually refer to as a *construct*.  For example, let's say that I want to test how well you understand the distinction between the different types of numbers described above.  I could give you a pop quiz that would ask you several questions about these concepts and count how many you got right.  This test might or might not be a good measurement of the construct of your actual knowledge -- for example, if I were to write the test in a confusing way or use language that you don't understand, then the test might suggest you don't understand the concepts when really you do. On the other hand, if I give a multiple choice test with very obvious wrong answers, then you might be able to perform well on the test even if you don't actually understand the material.-->  
En muchas áreas, como en la psicología, aquello que estamos midiendo no es una característica física, sino más bien un concepto teórico inobservable, a lo cual usualmente nos referimos como un *constructo*. Por ejemplo, digamos que quiero probar qué tan bien entiendes la distinción entre los diferentes tipos de números descritos anteriormente. Te podría dar un examen sorpresa en donde te haría varias preguntas sobre estos conceptos y contaría cuántas respuestas tienes correctas. Esta prueba puede o puede no ser una buena medición del constructo de tu conocimiento real-- por ejemplo, si escribiera una prueba en una forma confusa o un lenguaje que tú no entiendes, entonces la prueba puede sugerir que no entiendes los conceptos cuando en realidad sí los entiendes. Por otro lado, si te doy una prueba de opción múltiple con muchas respuestas obviamente incorrectas, entonces es posible que puedas desempeñarte bien en la prueba, incluso si en realidad no comprendes el material. 

<!--It is usually impossible to measure a construct without some amount of error.  In the example above, you might know the answer, but you might mis-read the question and get it wrong.  In other cases, there is error intrinsic to the thing being measured, such as when we measure how long it takes a person to respond on a simple reaction time test, which will vary from trial to trial for many reasons. We generally want our measurement error to be as low as possible, which we can achieve either by improving the quality of the measurement (for example, using a better time to measure reaction time), or by averaging over a larger number of indvidiual measurements.   -->
Usualmente es imposible medir un constructo sin cierto margen de error. En el ejemplo de arriba, puede que sepas la respuesta, pero puede que hayas leído mal la pregunta y por ende, obtenido una respuesta incorrecta. En otros casos, puede haber errores intrínsecos con respecto a aquella cosa que quiere ser medida, como cuando medimos cuánto le toma a una persona reaccionar en una simple prueba de tiempo de reacción, las cuales pueden variar de prueba en prueba por muchas razones. Generalmente queremos que nuestro error de medición sea lo más bajo posible, lo que podemos lograr mejorando la calidad de nuestras mediciones (por ejemplo, usando una mejor medición del tiempo para medir el tiempo de reacción), o promediando un mayor número de mediciones individuales.

<!--Sometimes there is a standard against which other measurements can be tested, which we might refer to as a "gold standard" -- for example, measurement of sleep can be done using many different devices (such as devices that measure movement in bed), but they are generally considered inferior to the gold standard of polysomnography (which uses measurement of brain waves to quantify the amount of time a person spends in each stage of sleep).  Often the gold standard is more difficult or expensive to perform, and the cheaper method is used even though it might have greater error.-->
A veces existe un estándar con el que se pueden probar otras mediciones, al que podríamos referirnos como un "estándar dorado" -- por ejemplo, la medición del sueño se puede realizar utilizando muchos dispositivos diferentes (como dispositivos que miden el movimiento de una persona mientras duerme), pero generalmente se consideran inferiores al estandar dorado de la polisomnografía (el cual es un examen que mide ondas cerebrales para cuantificar la cantidad de tiempo que una persona pasa en cada etapa del sueño). A menudo, el estandar dorado es más difícil o más caro de utilizar, y el método más barato es usado incluso cuando pueda tener un mayor margen de error. 

<!--When we think about what makes a good measurement, we usually distinguish two different aspects of a good measurement: it should be *reliable*, and it should be *valid*.-->
Cuando pensamos en aquello que constituye a una buena medición, usualmente distinguimos dos diferentes aspectos que debe tener: debe de ser *confiable*, y debe de ser *válida*. 

<!--### Reliability-->
### Confiabilidad

<!--Reliability refers to the consistency of our measurements.  One common form of reliability, known as "test-retest reliability", measures how well the measurements agree if the same measurement is performed twice.  For example, I might give you a questionnaire about your attitude towards statistics today, repeat this same questionnaire tomorrow, and compare your answers on the two days; we would hope that they would be very similar to one another, unless something happened in between the two tests that should have changed your view of statistics (like reading this book!).-->  
La confiabilidad se refiere a la consistencia de nuestras mediciones. Una forma común de confiabilidad, conocida como "confiabilidad test-retest", mide qué tan bien concuerdan las mediciones si la misma medición se realiza dos veces. Por ejemplo, si te doy un cuestionario sobre tu actitud con respecto a la estadística hoy, y repito este mismo cuestionario mañana, al comparar tus respuestas en los dos días esperaríamos que tuvieran resultados muy similares entre sí, a menos que algo sucediera entre la aplicación de ambos cuestionarios que haya cambiado tu perspectiva de la estadística (¡como leer este libro!).

<!--Another way to assess reliability comes in cases where the data include subjective judgments.  For example, let's say that a researcher wants to determine whether a treatment changes how well an autistic child interacts with other children, which is measured by having experts watch the child and rate their interactions with the other children.  In this case we would like to make sure that the answers don't depend on the individual rater --- that is, we would like for there to be high *inter-rater reliability*.  This can be assessed by having more than one rater perform the rating, and then comparing their ratings to make sure that they agree well with one another.-->
Otra forma de evaluar la confiabilidad surge en casos en que los datos incluyen juicios subjetivos. Por ejemplo, digamos que unx investigadorx quiere determinar si un tratamiento cambia qué tan bien interactúa unx niñx que se encuentra dentro del espectro autista con otros niñxs, lo cual es medido a través de expertos que observan al niñx y califican sus interacciones con lxs otrxs niñxs. En este caso queremos asegurarnos de que las respuestas no dependan del individuo que está calificando-- nos gustaría que existiera una alta *confiabilidad entre calificadores*. Esto puede ser evaluado teniendo más de unx solx evaluadorx, y después al comparar sus calificaciones asegurarnos de que concuerden entre sí.   

<!--Reliability is important if we want to compare one measurement to another, because the relationship between two different variables can't be any stronger than the relationship between either of the variables and itself (i.e., its reliability).  This means that an unreliable measure can never have a strong statistical relationship with any other measure.  For this reason, researchers developing a new measurement (such as a new survey) will often go to great lengths to establish and improve its reliability.-->
La confiabilidad es importante si queremos comparar una medición con otra, porque la relación entre dos variables diferentes no puede ser más fuerte que la relación entre cualquiera de las variables y ella misma (es decir, su confiabilidad). Esto significa que una medición no confiable nunca puede tener una relación estadísticamente fuerte con cualquier otra medición. Por esta razón, lxs investigadorxs que desarrollan una nueva medición (como un nuevo cuestionario) a menudo realizarán todo lo posible para establecer y mejorar su confiabilidad. 

```{r ReliabilityValidity, echo=FALSE,fig.cap="A figure demonstrating the distinction between reliability and validity, using shots at a bullseye. Reliability refers to the consistency of location of shots, and validity refers to the accuracy of the shots with respect to the center of the bullseye. ",fig.width=6,fig.height=6,out.height='33%'}
# Figure inspired by https://commons.wikimedia.org/wiki/File:Reliability_and_validity.svg

# from https://stackoverflow.com/questions/6862742/draw-a-circle-with-ggplot2

set.seed(12345)
gg_circle <- function(r, xc, yc, color="black", fill=NA, ...) {
    x <- xc + r*cos(seq(0, pi, length.out=100))
    ymax <- yc + r*sin(seq(0, pi, length.out=100))
    ymin <- yc + r*sin(seq(0, -pi, length.out=100))
    annotate("ribbon", x=x, ymin=ymin, ymax=ymax, color=color, fill=fill, ...)
}


npoints <- 16
rel_mult <- 0.75
unrel_mult <- 2.5

plotDf <- tibble(
  X_RelVal = rnorm(npoints)*rel_mult,
  Y_RelVal = rnorm(npoints)*rel_mult,
  X_RelInval = rnorm(npoints)*rel_mult+2,
  Y_RelInval = rnorm(npoints)*rel_mult+2,
  X_UnrelInval = rnorm(npoints)*unrel_mult+2,
  Y_UnrelInval = rnorm(npoints)*unrel_mult+2,
  X_UnrelVal = rnorm(npoints)*unrel_mult,
  Y_UnrelVal = rnorm(npoints)*unrel_mult
)

pointsize <- 3
linesize=2
p1=ggplot(plotDf,aes(X_RelVal,Y_RelVal)) +
  gg_circle(r=4, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=3, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=2, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=1, xc=0.0, yc=0.0,size=linesize,color='red',fill='red') + 
  geom_point(size=pointsize) + 
  xlim(-10,10) + ylim(-10,10) + 
  theme_void()

p2=ggplot(plotDf,aes(X_UnrelVal,Y_UnrelVal)) +
  gg_circle(r=4, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=3, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=2, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=1, xc=0.0, yc=0.0,size=linesize,color='red',fill='red') +
  geom_point(size=pointsize) + 
  xlim(-10,10) + ylim(-10,10) + 
  theme_void() 

p3=ggplot(plotDf,aes(X_RelInval,Y_RelInval)) +
  gg_circle(r=4, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=3, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=2, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=1, xc=0.0, yc=0.0,size=linesize,color='red',fill='red') + 
  geom_point(size=pointsize) + 
  xlim(-10,10) + ylim(-10,10) + 
  theme_void() 

p4=ggplot(plotDf,aes(X_UnrelInval,Y_UnrelInval)) +
  gg_circle(r=4, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=3, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=2, xc=0.0, yc=0.0,size=linesize,color='red') + 
  gg_circle(r=1, xc=0.0, yc=0.0,size=linesize,color='red',fill='red') + 
  geom_point(size=pointsize) + 
  xlim(-10,10) + ylim(-10,10) + 
  theme_void() 

plot_grid(p1,p2,p3,p4,ncol=2,label_size=12,
          labels=c('A: Reliable and valid',
                   'B: Unreliable but valid',
                   'C: Reliable but invalid',
                   'D: Unreliable and invalid'))


```

<!--### Validity-->
### Validez

<!--Reliability is important, but on its own it's not enough: After all, I could create a perfectly reliable measurement on a personality test by re-coding every answer using the same number, regardless of how the person actually answers.  We want our measurements to also be *valid* --- that is, we want to make sure that we are actually measuring the construct that we think we are measuring (Figure \@ref(fig:ReliabilityValidity)). There are many different types of validity that are commonly discussed; we will focus on three of them.-->
La confiabilidad es importante, pero por sí misma no es suficiente: Después de todo, todo lo que pueda crear mediciones perfectamente confiables en una prueba de personalidad a través de re-codificar todas las respuestas utilizando el mismo número, a pesar de cómo responda la persona. Queremos que nuestras mediciones sean también *válidas*-- esto quiere decir que, nos queremos asegurar de que en realidad estemos midiendo el constructo que pensamos que estamos midiendo (Figura \@ref(fig:ReliabilityValidity)). Existen varios tipos diferentes de validez que son comúnmente discutidos; a continuación nos enfocaremos en tres de ellos. 

<!--*Face validity*. Does the measurement make sense on its face?  If I were to tell you that I was going to measure a person's blood pressure by looking at the color of their tongue, you would probably think that this was not a valid measure on its face.  On the other hand, using a blood pressure cuff would have face validity.  This is usually a first reality check before we dive into more complicated aspects of validity.-->
*Validez aparente*. ¿La medición tiene sentido de forma aparente? Si te dijera que voy a medir la presión sanguínea de una persona con sólo observar el color de su lengua, probablemente pensarías que esta no es una medición válida aparente. Por otro lado, al utilizar un brazalete para medir la presión sanguínea tiene validez aparente. Esto es solamente un ejemplo simple antes de que nos centremos en aspectos más complejos de la validez.  

<!--*Construct validity*.  Is the measurement related to other measurements in an appropriate way?  This is often subdivided into two aspects.  *Convergent validity* means that the measurement should be closely related to other measures that are thought to reflect the same construct. Let's say that I am interested in measuring how extroverted a person is using either a questionnaire or an interview.  Convergent validity would be demonstrated if both of these different measurements are closely related to one another.  On the other hand, measurements thought to reflect different constructs should be unrelated, known as *divergent validity*.  If my theory of personality says that extraversion and conscientiousness are two distinct constructs, then I should also see that my measurements of extraversion are *unrelated* to measurements of conscientiousness.-->  
*Validez de constructo*. Primero hay que preguntarnos, ¿es esta medición relacionada con otras mediciones de una forma apropiada? A menudo esto se subdivide en dos aspectos. *Validez convergente* quiere decir que la medición debería de estar estrechamente relacionada con otras mediciones que se supone reflejan el mismo constructo. Digamos que me interesa medir qué tan extrovertida es una persona mediante un cuestionario o una entrevista. La validez convergente se demostraría si estas dos medidas diferentes estuvieran estrechamente relacionadas entre sí. Por otro lado, las mediciones que se cree que reflejan diferentes constructos no deben estar relacionadas, lo que se conoce como *validez divergente*. Si mi teoría de la personalidad dice que la extraversión y la responsabilidad son dos constructos distintos, entonces también debería poder observar que mi medición de la extraversión *no está relacionada* con la medición de la responsabilidad.

<!--*Predictive validity*.  If our measurements are truly valid, then they should also be predictive of other outcomes.  For example, let's say that we think that the psychological trait of sensation seeking (the desire for new experiences) is related to risk taking in the real world.  To test for predictive validity of a measurement of sensation seeking, we would test how well scores on the test predict scores on a different survey that measures real-world risk taking.-->
*Validez predictiva* Si nuestras mediciones son verdaderamente válidas, entonces también deberían de poder predecir otros resultados. Por ejemplo, digamos que pensamos que el rasgo psicológico de la búsqueda de sensaciones (el deseo de nuevas experiencias) está relacionado con la toma de riesgos en el mundo real. Para probar la validez predictiva de una medición de la búsqueda de sensaciones, probaríamos qué tan bien los puntajes en la prueba predicen los puntajes en un cuestionario diferente que mide la toma de riesgos en el mundo real. 

<!--## Learning Objectives-->
<!--Having read this chapter, you should be able to:-->
<!-- * Distinguish between different types of variables (quantitative/qualitative, binary/integer/real, discrete/continuous) and give examples of each of these kinds of variables-->
<!-- * Distinguish between the concepts of reliability and validity and apply each concept to a particular dataset -->

## Objetivos de aprendizaje

Al haber leído este capítulo deberías de ser capaz de:

* Distinguir entre diferentes tipos de variables (cuantitativas/cualitativas, binarios/enteros/reales, discretos/continuos) y poder dar ejemplos de cada una de estas variables.
* Distinguir entre conceptos de confiabilidad y validez y poder aplicar cada concepto a un conjunto de datos en particular.

<!--## Suggested readings-->
## Lecturas sugeridas

- [*An Introduction to Psychometric Theory with Applications in R*](http://www.personality-project.org/r/book/) - A free online textbook on psychological measurement

<!--## Appendix-->
## Apéndice

<!--### Scales of measurement-->
### Escalas de medición
<!--All variables must take on at least two different possible values (otherwise they would be a *constant* rather than a variable), but different values of the variable can relate to each other in different ways, which we refer to as *scales of measurement*.  There are four ways in which the different values of a variable can differ.-->
Todas las variables deben tomar al menos dos valores diferentes posibles (de lo contrario, serían una *constante* en lugar de una variable), pero diferentes valores de la variable pueden relacionarse entre sí de diferentes maneras, a estas nos referimos como *escalas de medición*. Hay cuatro formas en las que pueden diferir los diferentes valores de una variable. 

<!-- - *Identity*: Each value of the variable has a unique meaning.-->
- *Identidad*: Cada valor de la variable tiene un significado único.
<!-- - *Magnitude*: The values of the variable reflect different magnitudes and have an ordered relationship to one another -- that is, some values are larger and some are smaller.-->
- *Magnitud*: Los valores de la variable reflejan diferentes magnitudes y tienen una relación ordenada entre sí-- por lo tanto, algunos valores son mayores y otros son menores.
<!-- - *Equal intervals*: Units along the scale of measurement are equal to one another. This means, for example, that the difference between 1 and 2 would be equal in its magnitude to the difference between 19 and 20.-->
- *Intervalos iguales*: Las unidades a lo largo de la escala de medición son iguales entre sí. Esto quiere decir, por ejemplo, que la diferencia entre 1 y 2 sería igual en su magnitud a la diferencia entre 19 y 20. 
<!-- - *Absolute zero*:  The scale has a true meaningful zero point. For example, for many measurements of physical quantities such as height or weight, this is the complete absence of the thing being measured.-->
- *Cero absoluto*: La escala tiene un verdadero punto cero significativo. Por ejemplo, para muchas mediciones de cantidades físicas como la altura o el peso, esta es la ausencia total de la cosa que está siendo medida. 

<!--There are four different scales of measurement that go along with these different ways that values of a variable can differ.-->
Hay cuatro escalas diferentes de medición que van de la mano con estas diferentes formas en que los valores de una variable pueden diferir. 

<!--*Nominal scale*.  A nominal variable satisfies the criterion of identity, such that each value of the variable represents something different, but the numbers simply serve as qualitative labels as discussed above.  For example, we might ask people for their political party affiliation, and then code those as numbers: 1 = "Republican", 2 = "Democrat", 3 = "Libertarian", and so on. However, the different numbers do not have any ordered relationship with one another.-->  
*Escala Nominal*. Una variable nominal satisface el criterio de identidad, de modo que cada valor de la variable representa algo diferente, pero los números simplemente sirven como etiquetas cualitativas, como mencionamos al principio. Por ejemplo, es posible que le preguntemos a las personas el partido politico al que se suscriben, y después codificar esa información como números: 1= "Republicanos", 2= "Demócrata", 3= "Libertaria", etc. Sin embargo, los números no tienen ninguna relación ordenada entre sí.

<!--*Ordinal scale*. An ordinal variable satisfies the criteria of identity and magnitude, such that the values can be ordered in terms of their magnitude.  For example, we might ask a person with chronic pain to complete a form every day assessing how bad their pain is, using a 1-7 numeric scale. Note that while the person is presumably feeling more pain on a day when they report a 6 versus a day when they report a 3, it wouldn't make sense to say that their pain is twice as bad on the former versus the latter day; the ordering gives us information about relative magnitude, but the differences between values are not necessarily equal in magnitude.-->
*Escala ordinal*. Una variable ordinal satisface el criterio de identidad y magnitud, como que el valor puede ser ordenado en términos de su magnitud. Por ejemplo, le podemos preguntar a una persona con dolor crónico que llene un formato diario en donde evalúe qué tan mal siente su dolor, utilizando una escala numérica del 1 al 7. Hay que tomar en cuenta que, si bien la persona presumiblemente siente más dolor en un día en el que reporta un 6 frente a un día en que reporta un 3, no tendría sentido decir que su dolor es dos veces más intenso en el primero que en el último día; el orden nos da información sobre la magnitud relativa, pero las diferencias entre los valores no son necesariamente iguales en magnitud. 

<!--*Interval scale*. An interval scale has all of the features of an ordinal scale, but in addition the intervals between units on the measurement scale can be treated as equal.  A standard example is physical temperature measured in Celsius or Fahrenheit; the physical difference between 10 and 20 degrees is the same as the physical difference between 90 and 100 degrees, but each scale can also take on negative values.-->
*Escala de Intervalo*. Una escala de intervalo tiene todas las características de una escala ordinal, pero además los intervalos entre unidades en la escala de medición pueden tratarse como iguales. Un ejemplo estándar es la temperatura física medida en grados Celsius o Fahrenheit; la diferencia física entre 10 y 20 grados es la misma que la diferencia física entre 90 y 100 grados, pero cada escala también puede tomar valores negativos.

<!--*Ratio scale*. A ratio scale variable has all four of the features outlined above: identity, magnitude, equal intervals, and absolute zero.  The difference between a ratio scale variable and an interval scale variable is that the ratio scale variable has a true zero point.  Examples of ratio scale variables include physical height and weight, along with temperature measured in Kelvin.-->
*Escala de proporción (o de razón)*. Una variable a escala de proporción/razón tiene las cuatro características que se describen anteriormente: Identidad, magnitud, intervalos iguales y cero absoluto. La diferencia entre una variable de escala de razón y una variable de escala de intervalo es que la variable de escala de razón tiene un verdadero punto cero. Ejemplos de variables de escala de razón incluyen la altura y el peso físicos, junto con la temperatura medida en Kelvin.

<!--There are two important reasons that we must pay attention to the scale of measurement of a variable.  First, the scale determines what kind of mathematical operations we can apply to the data (see Table \@ref(tab:MeasurementTypes)).  A nominal variable can only be compared for equality; that is, do two observations on that variable have the same numeric value?  It would not make sense to apply other mathematical operations to a nominal variable, since they don't really function as numbers in a nominal variable, but rather as labels.  With ordinal variables, we can also test whether one value is greater or lesser than another, but we can't do any arithmetic.  Interval and ratio variables allow us to perform arithmetic; with interval variables we can only add or subtract values, whereas with ratio variables we can also multiply and divide values.-->
Hay dos razones importantes a las cuales les debemos de prestar atención a la escala de medición de la variable. En primer lugar, la escala determina qué tipo de operaciones matemáticas podemos aplicar a los datos (see Table \@ref(tab:MeasurementTypes)). Una variable nominal solamente se puede comparar por igualdad; es decir, ¿dos observaciones de esa variable tienen el mismo valor numérico? No tendría sentido aplicar otras operaciones matemáticas a una variable nominal, ya que en realidad no funcionan como números en una variable nominal, sino más bien como etiquetas. Con las variables ordinales, también podemos probar si un valor es mayor o menor que otro, pero no podemos hacer ninguna aritmética. Las variables de intervalo y razón nos permiten realizar operaciones aritméticas; con variables de intervalo solo podemos sumar o restar valores, mientras que con variables de razón también podemos multiplicar y dividir valores.

```{r echo=FALSE,MeasurementTypes}

measTypes=data.frame(equal=c('OK','OK','OK','OK'),
                     gtlt=c('','OK','OK','OK'),
                     addsub=c('','','OK','OK'),
                     multdiv=c('','','','OK'),
                     row.names=c('Nominal','Ordinal','Interval','Ratio'))
knitr::kable(
  measTypes, booktabs = TRUE,
  caption = 'Different scales of measurement admit different types of numeric operations',
  col.names=c("Equal/not equal",">/<", "+/-", "Multiply/divide")
)


```

<!--These constraints also imply that there are certain kinds of statistics that we can compute on each type of variable.  Statistics that simply involve counting of different values (such as the most common value, known as the *mode*), can be calculated on any of the variable types.  Other statistics are based on ordering or ranking of values (such as the *median*, which is the middle value when all of the values are ordered by their magnitude), and these require that the value at least be on an ordinal scale.  Finally, statistics that involve adding up values (such as the average, or *mean*), require that the variables be at least on an interval scale.  Having said that, we should note that it's quite common for researchers to compute the mean of variables that are only ordinal (such as responses on personality tests), but this can sometimes be problematic.-->
Estas restricciones también implican que existen ciertos tipos de estadística que podemos calcular sobre cada tipo de variable. La estadística que solamente se trate de contar los diferentes valores (como el valor más común comunido como *modo*), puede ser calculado en cualquiera de los tipos de variables. Otro tipo de estadística está basada en ordenar o en clasificar los valores (como la *mediana*, la cual es el valor que está en medio cuando todos los valores son ordenados por su magnitud), y estos requieren que el valor al menos esté en una escala ordinal. Finalmente, la estadística que se encarga de sumar los valores (como el promedio o *media*), requiere que las variables sean al menos en una escala de intervalo. Habiendo dicho esto, debemos tomar en cuenta que es común que lxs investigadorxs calculen la media de variables que son solo ordinales (como las respuestas en las pruebas de personalidad), pero esto a veces puede ser problemático.






